High-level cognition is supported by information-rich but compressible brain activity patterns

### 2023-12-22 paper/main.tex
working on addressing R2 and R3 concerns.
i've also added some notes on the network shuffling/permuation analysis, which still needs to be implemented and integrated into the paper... [WIP]

We used a previously collected neuroimaging dataset to estimate the extent to
which each of these two possibilities might hold. The dataset we examined
comprised functional magnetic resonance imaging (fMRI) data collected as
participants listened to an audio recording of a 10-minute story, temporally
scrambled recordings of the story, or underwent a resting state
scan~\citep{SimoEtal16}. Each of these experimental conditions evokes different
depths of cognitive processing~\citep{SimoEtal16,LernEtal11,
HassEtal08,OwenEtal21}. We used across-participant classifiers to decode
listening times in each condition, as a proxy for how ``informative'' the
task-specific activity patterns were~\citep{SimoChan20}. We also used principle
components analysis to generate lower-dimensional representations of the
activity patterns. We then repeated the classification analyses after
preserving different numbers of components and examined how classification
accuracy changed across the different experimental conditions.

@@

We analyzed a dataset collected by \cite{SimoEtal16} that comprised four
experimental conditions. These conditions exposed participants to stimuli that
systematically varied in cognitive engagement. In the \textit{intact}
experimental condition, participants listened to an audio recording of a
10-minute Moth Radio Hour story, \textit{Pie Man}, by Jim O'Grady. In the
\textit{paragraph}-scrambled experimental condition, participants listened to a
temporally scrambled version of the story, where the paragraphs occurred out of
order, but where the same set of paragraphs was presented over the entire
listening interval. All participants in this condition experienced the
scrambled paragraphs in the same order. In the \textit{word}-scrambled
experimental condition, participants listened to a temporally scrambled version
of the story, where the words occurred in a random order. Again, all
participants in this condition experienced the scrambled words in the same
order. Finally, in the \textit{rest} experimental condition, participants lay
in the scanner with no overt stimulus, while keeping their eyes open and
blinking as needed. This public dataset provided a convenient means for testing
our hypothesis that different levels of cognitive processing and engagement
affect how informative and compressible the associated brain patterns are.

@@

If the informativeness and compressibility of brain activity patterns vary over
time, do these properties might also vary across brain networks? We used a
network parcellation identified by \cite{YeoEtal11} to segment the brain into
seven distinct networks. The networks can be sorted (roughly) in order from
lower-level to higher-level cortex as follows (Figs.~\ref{fig:networks}A--C):
visual, somatomotor, dorsal attention, ventral attention, limbic,
frontoparietal, and default mode. Next, we computed decoding curves separately
for the activity patterns within each network and identified each network's
inflection point, for each experimental condition. Moving from low-order
networks to higher-order networks, we found that decoding accuracy tended to
increase in the higher-level experimental conditions and decrease (slightly) in
the lower-level experimental conditions (Fig.~\ref{fig:networks}D, E;
Spearman's rank correlation between decoding accuracy and network order:
intact: $\rho = 0.362, p < 0.001$; paragraph: $\rho = 0.441, p < 0.001$; word:
$\rho = -0.102, p = 0.007$; rest: $\rho = -0.354, p < 0.001$). This suggests
that higher-order networks may carry more content-relevant or stimulus-driven
``information.'' We found no clear trends in the proportions of components
required to achieve 5\% decoding accuracy across networks or conditions
(Fig.~\ref{fig:networks}F).

@@

We examined an fMRI dataset collected by \cite{SimoEtal16} that the authors
have made publicly available at
\href{http://arks.princeton.edu/ark:/88435/dsp015d86p269k}{arks.princeton.edu/ark:/88435/dsp015d86p269k}.
The dataset comprises neuroimaging data collected as participants listened to
an audio recording of a story (intact condition; 36 participants), listened to
temporally scrambled recordings of the same story (17 participants in the
paragraph-scrambled condition listened to the paragraphs in a randomized order
and 36 in the word-scrambled condition listened to the words in a randomized
order), or lay resting with their eyes open in the scanner (rest condition; 36
participants). Full neuroimaging details may be found in the original paper for
which the data were collected~\citep{SimoEtal16}. Procedures were approved by
the Princeton University Committee on Activities Involving Human Subjects, and
by the Western Institutional Review Board (Puyallup, WA). All subjects were
native English speakers with normal hearing and provided written informed
consent.

@@

>>>>

We used a previously collected neuroimaging dataset to estimate the extent to
which each of these two possibilities might hold. The dataset we examined
comprised functional magnetic resonance imaging (fMRI) data collected as
participants listened to an audio recording of a 7-minute story, temporally
scrambled recordings of the story, or underwent a resting state
scan~\citep{SimoEtal16}. Each of these experimental conditions evokes different
depths of cognitive processing~\citep{SimoEtal16,LernEtal11,
HassEtal08,OwenEtal21}. We used across-participant classifiers to decode
listening times in each condition, as a proxy for how ``informative'' the
task-specific activity patterns were~\citep{SimoChan20}. We also used principle
components analysis to generate lower-dimensional representations of the
activity patterns. We then repeated the classification analyses after
preserving different numbers of components and examined how classification
accuracy changed across the different experimental conditions.

@@

We analyzed a dataset collected by \cite{SimoEtal16} that comprised four
experimental conditions. These conditions exposed participants to stimuli that
systematically varied in cognitive engagement. In the \textit{intact}
experimental condition, participants listened to an audio recording of a
7-minute Moth Radio Hour story, \textit{Pie Man}, by Jim O'Grady. In the
\textit{paragraph}-scrambled experimental condition, participants listened to a
temporally scrambled version of the story, where the paragraphs occurred out of
order, but where the same set of paragraphs was presented over the entire
listening interval. All participants in this condition experienced the
scrambled paragraphs in the same order. In the \textit{word}-scrambled
experimental condition, participants listened to a temporally scrambled version
of the story, where the words occurred in a random order. Again, all
participants in this condition experienced the scrambled words in the same
order. Finally, in the \textit{rest} experimental condition, participants lay
in the scanner with no overt stimulus, while keeping their eyes open and
blinking as needed. This public dataset provided a convenient means for testing
our hypothesis that different levels of cognitive processing and engagement
affect how informative and compressible the associated brain patterns are.

@@

If the informativeness and compressibility of brain activity patterns vary over
time, do these properties might also vary across brain networks? We used a
network parcellation identified by \cite{YeoEtal11} to segment the brain into
seven distinct networks. The networks can be sorted (roughly) in order from
lower-level to higher-level cortex as follows (Figs.~\ref{fig:networks}A--C):
visual, somatomotor, dorsal attention, ventral attention, limbic,
frontoparietal, and default mode. Next, we computed decoding curves separately
for the activity patterns within each network and identified each network's
inflection point, for each experimental condition. Moving from low-order
networks to higher-order networks, we found that decoding accuracy tended to
increase in the higher-level experimental conditions and decrease (slightly) in
the lower-level experimental conditions (Fig.~\ref{fig:networks}D, E;
Spearman's rank correlation between decoding accuracy and network order:
intact: $\rho = 0.362, p < 0.001$; paragraph: $\rho = 0.441, p < 0.001$; word:
$\rho = -0.102, p = 0.007$; rest: $\rho = -0.354, p < 0.001$). This suggests
that higher-order networks may carry more content-relevant or stimulus-driven
``information.'' We found no clear trends in the proportions of components
required to achieve 5\% decoding accuracy across networks or conditions
(Fig.~\ref{fig:networks}F). We note that the limbic network we considered here
often overlaps with low (imaging) signal regions, and therefore it may be
difficult to draw strong conclusions about this network's informativeness or
compressibility.

@@

We examined an fMRI dataset collected by \cite{SimoEtal16} that the authors
have made publicly available at
\href{http://arks.princeton.edu/ark:/88435/dsp015d86p269k}{arks.princeton.edu/ark:/88435/dsp015d86p269k}.
The dataset comprises neuroimaging data collected as participants listened to
an audio recording of a story (intact condition; 36 participants), listened to
temporally scrambled recordings of the same story (17 participants in the
paragraph-scrambled condition listened to the paragraphs in a randomized order
and 36 in the word-scrambled condition listened to the words in a randomized
order), or lay resting with their eyes open in the scanner (rest condition; 36
participants). Full neuroimaging details may be found in the original paper for
which the data were collected~\citep{SimoEtal16}. Procedures were approved by
the Princeton University Committee on Activities Involving Human Subjects, and
by the Western Institutional Review Board (Puyallup, WA). All subjects were
native English speakers with normal hearing and provided written informed
consent. We have excerpted the relevant portions of the dataset documention
here to provide information about the scanning parameters and preprocessing
steps used to generate the data we analyzed (the original descriptions may be
found at the above link):

\begin{quote}
Subjects were scanned in a 3T full-body MRI scanner (Skyra; Siemens) with a
sixteen-channel head coil. For functional scans, images were acquired using a
T2* weighted echo planer imaging (EPI) pulse sequence [repetition time (TR),
1500 ms; echo time (TE), 28 ms; flip angle, $64^\circ$], each volume comprising
27 slices of 4 mm thickness with 0 mm gap; slice acquisition order was
interleaved. In-plane resolution was $3 \times 3$ mm\textsuperscript{2} [field
of view (FOV), 192 $\times$ 192 mm\textsuperscript{2}]. Anatomical images were
acquired using a T1-weighted magnetization-prepared rapid-acquisition gradient
echo (MPRAGE) pulse sequence (TR, 2300 ms; TE, 3.08 ms; flip angle 9$^\circ$;
0.89 mm\textsuperscript{3} resolution; FOV, 256 mm\textsuperscript{2}). To
minimize head movement, subjects' heads were stabilized with foam padding.
Stimuli were presented using the Psychophysics toolbox~\citep{Brai97,Pell97}.
Subjects were provided with an MRI compatible in-ear mono earbuds (Sensimetrics
Model S14), which provided the same audio input to each ear. MRI-safe passive
noise-canceling headphones were placed over the earbuds, for noise removal and
safety.

Functional data were preprocessed and analyzed using FSL
(www.fmrib.ox.ac.uk/fsl), including correction for head motion and
slice-acquisition time, spatial smoothing (6 mm FWHM Gaussian kernel), and
high-pass temporal filtering (140 s period). Preprocessed data were aligned to
coplanar and high-resolution anatomicals and the standard MNI152 brain, and
interpolated to 3-mm isotropic voxels.
\end{quote}

The intact and word conditions each comprised 300 TRs (7.5 minutes) per participant.
The paragraph condition comprised 272 TRs (6.8 minutes) per participant. The rest
condition comprised 400 TRs (10 minutes) per participant.

@@

We also considered alternative approaches to obtaining compact representations
of the neuroimaging data, including network
parcellations~\citep[e.g.,][]{SchaEtal18, GordEtal16}. Whereas network
parcellations are typically derived from large resting state datasets, HTFA may
be applied to much smaller datasets. In our prior work, we showed that HTFA
applied to the same dataset used here can explain full-brain activity to within
a maximum of 0.25 standard deviations of each voxel's observed activity in the
original dataset, taken across all voxels, images, and participants, using the
700-node representation we also employed here~\citep{MannEtal18}. Some of the
explanatory power of HTFA comes from the fact that each node's explanatory
power falls off smoothly with distance to its center. Intuitively, the result
is a representation that looks like a lightly spatially smoothed version of the
original data, but where the degree of smoothing varies across the brain
according to how spatially autocorrelated the local activity patterns are.
